---
title: "Getting started with eavsr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with eavsr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`eavsr` provides a tidy interface to the U.S. Election Assistance Commission's
**Election Administration and Voting Survey (EAVS)** — the largest public
dataset on local election administration in the United States. This vignette
walks through the main workflow: discovering cycles, downloading and reading
data, harmonizing variables, classifying jurisdictions, and rolling everything
up to county-equivalent geography.

> Not affiliated with or endorsed by the U.S. Election Assistance Commission.

```{r setup}
library(eavsr)
```

## Which cycles are available?

`eavs_cycles()` returns the cycles the package knows about, the source URLs,
and whether each is ready to download (`supported`) or still being verified
(`planned`).

```{r}
eavs_cycles()
```

## Downloading and reading data

`read_eavs()` downloads a cycle (if it is not already cached), reads it into a
tibble, and recodes EAVS missing-value codes (negative integers such as `-99`)
to `NA`. The download happens once and is cached locally, so later calls are
fast.

```{r, eval = FALSE}
# One call: download (first time only), unzip, read, and clean.
eavs <- read_eavs(2024)

# The file exactly as released, with no recoding or name cleaning:
raw <- read_eavs(2024, level = "raw")
```

`read_eavs(level = "jurisdiction")` (the default) lower-cases column names with
`janitor::clean_names()` and recodes missing values; `level = "raw"` returns the
file untouched. You can also call `download_eavs()` directly if you only want the
file path.

### The codebook

EAVS ships a codebook describing all ~500 variables. For cycles whose codebook
is released as a spreadsheet, `eavs_codebook()` returns it as a tibble (this
requires the `readxl` package):

```{r, eval = FALSE}
cb <- eavs_codebook(2024)
```

## Harmonizing variables across cycles

EAVS variable codes drift across survey cycles, which makes multi-cycle work
painful. `eavs_crosswalk()` returns a hand-curated map from raw EAVS codes to a
stable harmonized name and topical section:

```{r}
eavs_crosswalk()
```

## Understanding EAVS geography

EAVS reporting units vary by state, and — importantly — so does the `FIPSCode`
field. There are three patterns to know about:

- **Most states** use a 10-digit code. A trailing `00000` marks a
  **county-level** row (county-reporting states); any other ending marks a
  **sub-county** row (the New England township states: ME, MA, NH, VT, CT, RI).
- **Wisconsin** uses a 5-digit internal code that is *not* a Census FIPS code;
  its ~1,850 municipalities carry the county only in the jurisdiction *name*
  (e.g. `TOWN OF ABRAMS - OCONTO COUNTY`).
- **Connecticut** reports by town and is mapped to its nine planning regions.

`classify_jurisdiction()` reads these patterns and adds `jurisdiction_type`,
`state_fips`, `county_fips`, and `is_multi_county`. It keys off the
`state_abbr` column (not the raw FIPS prefix), and applies Connecticut and
Wisconsin name lookups where needed.

```{r}
jur <- tibble::tibble(
  fips_code         = c("0100100000",     "2500112345",   "0900112345"),
  jurisdiction_name = c("AUTAUGA COUNTY", "EXAMPLE TOWN",  "ANDOVER TOWN"),
  state_abbr        = c("AL",             "MA",            "CT")
)

classify_jurisdiction(jur)
```

Here Autauga County (Alabama) is a county row; the Massachusetts town is a
sub-county row whose `county_fips` is derived from its code; and the Connecticut
town is mapped to its planning region (`09110`). The Connecticut and Wisconsin
lookup tables are extensible — pass your own via the `ct_lookup` and `wi_lookup`
arguments, or see `?ct_planning_regions` and `?wi_subcounty_fips`.

## Rolling up to county level

`aggregate_to_county()` produces one row per county-equivalent FIPS code. For
each county it prefers a directly reported county row; where only sub-county
rows exist, it sums their numeric columns up to the county. Ordinal questions
(like EAVS `d8`) can be summarized by their modal value instead of summed. It
calls `classify_jurisdiction()` for you if the data has not been classified.

```{r}
example <- tibble::tibble(
  fips_code         = c("0100100000",     "2500112345", "2500167890"),
  jurisdiction_name = c("AUTAUGA COUNTY", "TOWN A",      "TOWN B"),
  state_abbr        = c("AL",             "MA",          "MA"),
  total_voters      = c(40000,            12000,         8000)
)

aggregate_to_county(example)
```

The Alabama county is carried through directly (`data_source = "direct_county"`),
while the two Massachusetts towns are summed into their shared county
(`data_source = "aggregated_subcounty"`). By default U.S. territories and Alaska
are dropped, since they do not roll up cleanly to county FIPS; see
`?aggregate_to_county` to change that.

## Citing EAVS

The EAC asks researchers to cite the EAVS release. `cite_eavs()` builds the
recommended citation as text, BibTeX, or a structured list:

```{r}
cite_eavs(2024)
```

```{r}
cat(cite_eavs(2024, style = "bibtex"))
```

## Managing the cache

Downloaded files live under `tools::R_user_dir("eavsr", "cache")`. Override the
location with the `EAVSR_CACHE_DIR` environment variable.

```{r, eval = FALSE}
eavs_cache_dir()    # where files are stored
eavs_cache_info()   # what is currently cached
eavs_clear_cache()  # remove cached files
```

## Pairing with rurality data

EAVS county-level output pairs naturally with the
[`rurality`](https://github.com/cwimpy/rurality) package, which provides
county-level rurality classifications (RUCC, RUCA, IRR, OMB) for rural–urban
analyses of election administration.
